---
title: "W8_Genomics_Lab"
author: "Vera Sophia Beliaev"
date: "2/17/2022"
output: pdf_document
---


# Section 4: Population Analysis


> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.



```{r}
expr <-  read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
#number of samples
nrow(expr)
```


```{r}
#Make a table counting each genotype
table(expr$geno)
```

```{r}
summary(expr)
```


>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?
The A/A genotype has much higher expression than the G/G genotype as evidenced by the boxplot. Having an A/A genotype corresponds to high expression of ORMDL3 while having a G/G genotype corresponds to low expression of ORMDL3.

```{r}
library(ggplot2)
```

Make a boxplot

```{r}
ggplot(expr) + aes(geno, exp, col= geno) + geom_boxplot(notch=TRUE)
```













