---
title: "Grade Function Lab Week 5"
author: "Vera Sophia Beliaev"
date: "2/7/2022"
output: pdf_document
---

Today we are creating **R functions** to process grades of a class, dropping the lowest score for each person.

Questions:

> Q1. Write a function grade() to determine an overall grade from a vector of student homework
assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an
NA value) this can be used as a score to be potentially dropped. Your final function should be
adquately explained with code comments and be able to work on an example class gradebook
such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)

student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)

student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)

```

Write code for a simple version of the problem

```{r}
#Simple mean()
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
mean(student1)
```

Lowest score needs to be identified then dropped

```{r}
#Find which elemend of the vector contains the lowest score
which.min(student1)
```

Need to drop/exclude the lowest score from the mean()
```{r}
#This returns everything except the 8th element
student1[-8]

```

Integrate which.min to find everything except that element

```{r}
student1[-which.min(student1)]
```

Calc the mean of all the elements of the vect excluding which.min elem

```{r}
mean(student1[-which.min(student1)])
```

Looking to student 2 and 3

We try using the na.rm-TRUE arg but it's not fair
```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
mean(student2, na.rm=TRUE)


student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
mean(student3, na.rm=TRUE)
```

Replace all NA values with zero


Find the NA elements
```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
x <- student2
is.na(x)

#tell me which element contains NA value
which(is.na(x))
```

NA value is identified, now we need to switch it with a value of zero (mask them)

```{r}
x[is.na(x)]
x[is.na(x)] <- 0
x
mean(x)
```

We need to drop the lowest score
```{r}
x[is.na(x)] <- 0
mean(x[-which.min(x)])
```

Test of student 3
```{r}
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
x <- student3
x[is.na(x)] <- 0
mean(x[-which.min(x)])
```

The gold:
```{r}
x[is.na(x)] <- 0
mean(x[-which.min(x)])
```

## Now we make the function

snippet -> function

Function contains: a name, input args, the body


USing RStudio, select 'Code > Extract Function'
```{r}
grade <- function(x) {
  x[is.na(x)] <- 0
  mean(x[-which.min(x)])
}
```

```{r}
grade(student1)
grade(student2)
grade(student3)
```


```{r}
#' Calculate the average score for a vector of scores, dropping the lowest score. Missing values are given a value of zero.
#'
#' @param x A numeric vector of homework scores
#'
#' @return Average score
#' @export
#'
#' @examples
#' student <- c(100, NA, 90, 97)
#' grade(student)
#' 
grade <- function(x) {
  #Treat missing values as zero, mask NA with zero
   x[is.na(x)] <- 0
   #Exclude the lowest score from calculating the mean
  mean(x[-which.min(x)])
}
```

Now we apply the function to the whole class data
CSV format:"https://tinyurl.com/gradeinput"
```{r}
url <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names=1)
```

```{r}
apply(gradebook, 1, grade)
```



> Q2.  Using your grade() function and the supplied gradebook, Who is the top scoring student
overall in the gradebook? [3pts]

```{r}
#Find the highest scoring student on average

#through sorting
sort((apply(gradebook, 1, grade)))

#highest average score
max(apply(gradebook, 1, grade))

#which student scored highest
which.max(apply(gradebook, 1, grade))

```



> Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]

```{r}
gradebook
# Check stats of columns, check mean
which.min(apply(gradebook, 2, mean, na.rm=TRUE))

#Check stats of columns, check median
which.min(apply(gradebook, 2, median, na.rm=TRUE))

#make a boxplot
boxplot(gradebook)
 
```


> Q4. Optional Extension: From your analysis of the gradebook, which homework was most
predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

```{r}
#correlation b/w x & y (between average overall class score and average score for each homework), apply to the whole vector w/ apply(), 1st mask the NA values w/ zero
results <- apply(gradebook, 1, grade)


masked.gradebook <- gradebook
masked.gradebook[is.na(masked.gradebook)] <- 0


apply(masked.gradebook, 2, cor, x=results)


```


> Knit the doc to make a PDF using the dropdown menu or switch output at top to pdf_document











